Why do people punish? Evidence for a range of strategic concerns

### 2024-01-08 appendix.Rmd
prepare to submit pre-print

\setcounter{page}{1}
\centering
\noindent \hspace*{10mm} \small Why do people punish? Evidence for a range of strategic concerns \newline
\hspace*{1cm} \small Scott Claessens\textsuperscript{1}, Quentin D. Atkinson\textsuperscript{1}, Nichola Raihani\textsuperscript{1,2} \newline
\raggedright
\noindent \footnotesize \textsuperscript{1} School of Psychology, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{2} Department of Experimental Psychology, University College London, London, United Kingdom \newline
\normalsize
\newpage

@@

(ref:plotModel1aCaption) *Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model fitted to
the full dataset without pre-registered exclusions.* The model assumes an 
implementation error rate of 5%. Points represent posterior medians, line ranges 
represent 50% and 95% credible intervals. AI = advantageous inequity, DI = 
disadvantageous inequity.

@@

(ref:plotAllPers1Caption) *Posterior slopes from models including Big-6 
personality dimensions and Social Value Orientation, fitted to the full dataset 
without pre-registered exclusions.* Each row represents a separate model. Points 
represent posterior medians, line ranges represent 95% credible intervals. AI = 
advantageous inequity, DI = disadvantageous inequity.

@@

(ref:plotAllPolRel1Caption) *Posterior slopes from models including political 
ideology, views about social inequality, and religiosity, fitted to the full 
dataset without pre-registered exclusions.* Each row represents a separate model 
aside from Social Dominance Orientation and Right Wing Authoritarianism, which 
control for one another within the same model. Points represent posterior 
medians, line ranges represent 95% credible intervals. AI = advantageous 
inequity, DI = disadvantageous inequity.

@@

(ref:plotAllSliders1Caption) *Posterior slopes from models including 
self-reported strategy usage, fitted to the full dataset without pre-registered 
exclusions.* Each row represents a separate model. Highlighted estimates 
represent combinations where the self-report slider matched the behavioural 
strategy. Each strategy had an associated self-report slider except for the 
competitive strategy. Points represent posterior medians, line ranges represent 
95% credible intervals. AI = advantageous inequity, DI = disadvantageous 
inequity.

@@

(ref:plotTraceCaption) *Trace plots for parameter values from the Bayesian
latent state model fitted to data with exclusions.*

>>>>

\setcounter{page}{1}
\centering
\noindent \hspace*{10mm} \small Why do people punish? Evidence for a range of strategic concerns \newline
\hspace*{1cm} \small Scott Claessens\textsuperscript{1}, Quentin D. Atkinson\textsuperscript{1}, & Nichola Raihani\textsuperscript{1,2} \newline
\raggedright
\noindent \footnotesize \textsuperscript{1} School of Psychology, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{2} Department of Experimental Psychology, University College London, London, United Kingdom \newline
\par
\noindent{\bfseries{Contents}}
\contentsline {subsection}{Supplementary Figures S1-S13}{2}{}
\contentsline {subsection}{Supplementary Tables S1-S4}{15}{}
\normalsize
\newpage

@@

(ref:plotModel1aCaption) *Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model fitted to
the full dataset without pre-registered exclusions.* The model assumes an 
implementation error rate of 5%. Figure \@ref(fig:plotModel1b) in the main text
shows the same result, but from a model fitted to the reduced dataset with 
pre-registered exclusions. Points represent posterior medians, line ranges 
represent 50% and 95% credible intervals. AI = advantageous inequity, DI = 
disadvantageous inequity.

@@

(ref:plotAllPers1Caption) *Posterior slopes from models including Big-6 
personality dimensions and Social Value Orientation, fitted to the full dataset 
without pre-registered exclusions.* Each row represents a separate model. Figure 
\@ref(fig:plotAllPers2) in the main text shows the same results, but from models
fitted to the reduced dataset with pre-registered exclusions. Points represent
posterior medians, line ranges represent 95% credible intervals. AI = 
advantageous inequity, DI = disadvantageous inequity.

@@

(ref:plotAllPolRel1Caption) *Posterior slopes from models including political 
ideology, views about social inequality, and religiosity, fitted to the full 
dataset without pre-registered exclusions.* Each row represents a separate model 
aside from Social Dominance Orientation and Right Wing Authoritarianism, which 
control for one another within the same model. Figure \@ref(fig:plotAllPolRel2)
in the main text shows the same results, but from models fitted to the reduced
dataset with pre-registered exclusions. Points represent posterior medians, line 
ranges represent 95% credible intervals. AI = advantageous inequity, DI = 
disadvantageous inequity.

@@

(ref:plotAllSliders1Caption) *Posterior slopes from models including 
self-reported strategy usage, fitted to the full dataset without pre-registered 
exclusions.* Each row represents a separate model. Highlighted estimates 
represent combinations where the self-report slider matched the behavioural 
strategy. Figure \@ref(fig:plotAllSliders2) in the main text shows the same 
results, but from models fitted to the reduced dataset with pre-registered 
exclusions. Points represent posterior medians, line ranges represent 95% 
credible intervals. AI = advantageous inequity, DI = disadvantageous inequity.

@@

(ref:plotTraceCaption) *MCMC trace plots for parameter values from the Bayesian
latent state model fitted to data with exclusions.* These trace plots suggest
that the different chains mixed well and the model converged normally.

### 2024-01-08 manuscript.Rmd
update manuscript
prepare to submit pre-print

(ref:plotGamesCaption) *Visual summary of the six economic games.* In all
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does 
nothing. Player 1 is then given the option to punish by paying a certain amount 
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off after stealing, (B) Player 2 creates equality by stealing, (C) 
the computer "decides" whether Player 2 steals, (D) the fee-fine ratio is 1:1, 
(E) Player 2 is better off after stealing, and (F) Player 2 steals instead from 
a third-party.

@@

(ref:tableStrategiesCaption) \scriptsize Summary of the different functions for 
punishment and the behavioural strategies they predict. *Games A-F are 
the games employed in the current study (see Methods for more details). In each 
of the six games, participants are given the opportunity to punish players who 
"steal" and those who do not, meaning that participants make twelve punishment 
decisions in total. Each behavioural strategy implies a unique pattern of 
punishment across all decisions. Green ticks reflect decisions to punish, red 
crosses reflect decisions to not punish. In column headers, payoffs at the first 
stage (above) and the second stage (below) are denoted as P1-P2 (or P2-P3 [P1] 
for Game F) where participants take the role of P1 and P2 is the target of 
punishment. AI = advantageous inequity, DI = disadvantageous inequity.*

>>>>

(ref:plotGamesCaption) *Visual summary of the six economic games.* In the
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does 
nothing. Player 1 is then given the option to punish by paying a certain amount 
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off after stealing, (B) Player 2 creates equality by stealing, (C) 
the computer "decides" whether Player 2 steals, (D) the fee-fine ratio is 1:1, 
(E) Player 2 is better off after stealing, and (F) Player 2 steals instead from 
a third-party.

@@

(ref:tableStrategiesCaption) \scriptsize Summary of the different functions for 
punishment and the behavioural strategies they predict. *Games A-F are 
the games employed in the current study (see Methods for more details). In each 
of the six games, participants are given the opportunity to punish players who 
"steal" and those who do not, meaning that participants make twelve punishment 
decisions in total. Each behavioural strategy implies a unique pattern of 
punishment across all decisions (see Methods for detailed explanation of 
strategies). Green ticks reflect decisions to punish, red crosses reflect 
decisions to not punish. In column headers, payoffs at the first stage (above) 
and the second stage (below) are denoted as P1-P2 (or P2-P3 [P1] for Game F) 
where participants take the role of P1 and P2 is the target of punishment. AI = 
advantageous inequity, DI = disadvantageous inequity.*

### 2024-01-08 manuscript.tex
prepare to submit pre-print

\begin{figure}
\includegraphics[width=1\linewidth]{images/games} \caption{\emph{Visual summary of the six economic games.} In all
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does
nothing. Player 1 is then given the option to punish by paying a certain amount
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off after stealing, (B) Player 2 creates equality by stealing, (C)
the computer ``decides'' whether Player 2 steals, (D) the fee-fine ratio is 1:1,
(E) Player 2 is better off after stealing, and (F) Player 2 steals instead from
a third-party.}\label{fig:plotGames}
\end{figure}

@@

punishment across all decisions. Green ticks reflect decisions to punish, red
crosses reflect decisions to not punish. In column headers, payoffs at the first
stage (above) and the second stage (below) are denoted as P1-P2 (or P2-P3 {[}P1{]}
for Game F) where participants take the role of P1 and P2 is the target of
punishment. AI = advantageous inequity, DI = disadvantageous inequity.}}\\

@@

\leavevmode\vadjust pre{\hypertarget{ref-Laurin2012}{}}%
\CSLLeftMargin{64. }%
\CSLRightInline{Laurin, K., Shariff, A. F., Henrich, J. \& Kay, A. C. \href{https://doi.org/10.1098/rspb.2012.0615}{Outsourcing punishment to god: Beliefs in divine control reduce earthly punishment}. \emph{Proceedings of the Royal Society B: Biological Sciences} \textbf{279}, 3272--3281 (2012).}

@@

\noindent \hspace*{10mm} \small Why do people punish? Evidence for a range of strategic concerns \newline
\hspace*{1cm} \small Scott Claessens\textsuperscript{1}, Quentin D. Atkinson\textsuperscript{1}, Nichola Raihani\textsuperscript{1,2} \newline

@@

\noindent \footnotesize \textsuperscript{1} School of Psychology, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{2} Department of Experimental Psychology, University College London, London, United Kingdom \newline
\normalsize
\newpage

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotModel1a-1.pdf}
\caption{\label{fig:plotModel1a}\emph{Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model fitted to
the full dataset without pre-registered exclusions.} The model assumes an
implementation error rate of 5\%. Points represent posterior medians, line ranges
represent 50\% and 95\% credible intervals. AI = advantageous inequity, DI =
disadvantageous inequity.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPers1-1.pdf}
\caption{\label{fig:plotAllPers1}\emph{Posterior slopes from models including Big-6
personality dimensions and Social Value Orientation, fitted to the full dataset
without pre-registered exclusions.} Each row represents a separate model. Points
represent posterior medians, line ranges represent 95\% credible intervals. AI =
advantageous inequity, DI = disadvantageous inequity.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPolRel1-1.pdf}
\caption{\label{fig:plotAllPolRel1}\emph{Posterior slopes from models including political
ideology, views about social inequality, and religiosity, fitted to the full
dataset without pre-registered exclusions.} Each row represents a separate model
aside from Social Dominance Orientation and Right Wing Authoritarianism, which
control for one another within the same model. Points represent posterior
medians, line ranges represent 95\% credible intervals. AI = advantageous
inequity, DI = disadvantageous inequity.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllSliders1-1.pdf}
\caption{\label{fig:plotAllSliders1}\emph{Posterior slopes from models including
self-reported strategy usage, fitted to the full dataset without pre-registered
exclusions.} Each row represents a separate model. Highlighted estimates
represent combinations where the self-report slider matched the behavioural
strategy. Each strategy had an associated self-report slider except for the
competitive strategy. Points represent posterior medians, line ranges represent
95\% credible intervals. AI = advantageous inequity, DI = disadvantageous
inequity.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotTrace-1.pdf}
\caption{\label{fig:plotTrace}\emph{Trace plots for parameter values from the Bayesian
latent state model fitted to data with exclusions.}}
\end{figure}

>>>>

\begin{figure}
\includegraphics[width=1\linewidth]{images/games} \caption{\emph{Visual summary of the six economic games.} In the
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does
nothing. Player 1 is then given the option to punish by paying a certain amount
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off after stealing, (B) Player 2 creates equality by stealing, (C)
the computer ``decides'' whether Player 2 steals, (D) the fee-fine ratio is 1:1,
(E) Player 2 is better off after stealing, and (F) Player 2 steals instead from
a third-party.}\label{fig:plotGames}
\end{figure}

@@

punishment across all decisions (see Methods for detailed explanation of
strategies). Green ticks reflect decisions to punish, red crosses reflect
decisions to not punish. In column headers, payoffs at the first stage (above)
and the second stage (below) are denoted as P1-P2 (or P2-P3 {[}P1{]} for Game F)
where participants take the role of P1 and P2 is the target of punishment. AI =
advantageous inequity, DI = disadvantageous inequity.}}\\

@@

\leavevmode\vadjust pre{\hypertarget{ref-Laurin2012}{}}%
\CSLLeftMargin{64. }%
\CSLRightInline{Laurin, K., Shariff, A. F., Henrich, J. \& Kay, A. C. \href{https://doi.org/10.1098/rspb.2012.0615}{Outsourcing punishment to {God}: Beliefs in divine control reduce earthly punishment}. \emph{Proceedings of the Royal Society B: Biological Sciences} \textbf{279}, 3272--3281 (2012).}

@@

\noindent \hspace*{10mm} \small Why do people punish? Evidence for a range of strategic concerns \newline
\hspace*{1cm} \small Scott Claessens\textsuperscript{1}, Quentin D. Atkinson\textsuperscript{1}, \& Nichola Raihani\textsuperscript{1,2} \newline

@@

\noindent \footnotesize \textsuperscript{1} School of Psychology, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{2} Department of Experimental Psychology, University College London, London, United Kingdom \newline

\par

\noindent{\bfseries{Contents}}
\contentsline {subsection}{Supplementary Figures S1-S13}{2}{}
\contentsline {subsection}{Supplementary Tables S1-S4}{15}{}
\normalsize
\newpage

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotModel1a-1.pdf}
\caption{\label{fig:plotModel1a}\emph{Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model fitted to
the full dataset without pre-registered exclusions.} The model assumes an
implementation error rate of 5\%. Figure \ref{fig:plotModel1b} in the main text
shows the same result, but from a model fitted to the reduced dataset with
pre-registered exclusions. Points represent posterior medians, line ranges
represent 50\% and 95\% credible intervals. AI = advantageous inequity, DI =
disadvantageous inequity.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPers1-1.pdf}
\caption{\label{fig:plotAllPers1}\emph{Posterior slopes from models including Big-6
personality dimensions and Social Value Orientation, fitted to the full dataset
without pre-registered exclusions.} Each row represents a separate model. Figure
\ref{fig:plotAllPers2} in the main text shows the same results, but from models
fitted to the reduced dataset with pre-registered exclusions. Points represent
posterior medians, line ranges represent 95\% credible intervals. AI =
advantageous inequity, DI = disadvantageous inequity.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllPolRel1-1.pdf}
\caption{\label{fig:plotAllPolRel1}\emph{Posterior slopes from models including political
ideology, views about social inequality, and religiosity, fitted to the full
dataset without pre-registered exclusions.} Each row represents a separate model
aside from Social Dominance Orientation and Right Wing Authoritarianism, which
control for one another within the same model. Figure \ref{fig:plotAllPolRel2}
in the main text shows the same results, but from models fitted to the reduced
dataset with pre-registered exclusions. Points represent posterior medians, line
ranges represent 95\% credible intervals. AI = advantageous inequity, DI =
disadvantageous inequity.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotAllSliders1-1.pdf}
\caption{\label{fig:plotAllSliders1}\emph{Posterior slopes from models including
self-reported strategy usage, fitted to the full dataset without pre-registered
exclusions.} Each row represents a separate model. Highlighted estimates
represent combinations where the self-report slider matched the behavioural
strategy. Figure \ref{fig:plotAllSliders2} in the main text shows the same
results, but from models fitted to the reduced dataset with pre-registered
exclusions. Points represent posterior medians, line ranges represent 95\%
credible intervals. AI = advantageous inequity, DI = disadvantageous inequity.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/plotTrace-1.pdf}
\caption{\label{fig:plotTrace}\emph{MCMC trace plots for parameter values from the Bayesian
latent state model fitted to data with exclusions.} These trace plots suggest
that the different chains mixed well and the model converged normally.}
\end{figure}
